---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 20)

library(tidyverse)
```

# Day 09: Marble Mania

## Puzzle 1


You talk to the Elves while you wait for your navigation system to initialize. To pass the time, they introduce you to their favorite marble game.

The Elves play this game by taking turns arranging the marbles in a *circle* according to very particular rules. The marbles are numbered starting with 0 and increasing by 1 until every marble has a number.

First, the marble numbered 0 is placed in the circle. At this point, while it contains only a single marble, it is still a circle: the marble is both clockwise from itself and counter-clockwise from itself. This marble is designated the *current marble*.

Then, each Elf takes a turn placing the *lowest-numbered remaining marble* into the circle between the marbles that are 1 and 2 marbles clockwise of the current marble. (When the circle is large enough, this means that there is one marble between the marble that was just placed and the current marble.) The marble that was just placed then becomes the *current marble*.

However, if the marble that is about to be placed has a number which is a multiple of 23, *something entirely different happens*. First, the current player keeps the marble they would have placed, adding it to their *score*. In addition, the marble 7 marbles *counter-clockwise* from the current marble is *removed* from the circle and also added to the current player's score. The marble located immediately *clockwise* of the marble that was removed becomes the new *current marble*.

For example, suppose there are 9 players. After the marble with value 0 is placed in the middle, each player (shown in square brackets) takes a turn. The result of each of those turns would produce circles of marbles like this, where clockwise is to the right and the resulting current marble is in parentheses:

```
[-] (0)
[1]  0 (1)
[2]  0 (2) 1 
[3]  0  2  1 (3)
[4]  0 (4) 2  1  3 
[5]  0  4  2 (5) 1  3 
[6]  0  4  2  5  1 (6) 3 
[7]  0  4  2  5  1  6  3 (7)
[8]  0 (8) 4  2  5  1  6  3  7 
[9]  0  8  4 (9) 2  5  1  6  3  7 
[1]  0  8  4  9  2(10) 5  1  6  3  7 
[2]  0  8  4  9  2 10  5(11) 1  6  3  7 
[3]  0  8  4  9  2 10  5 11  1(12) 6  3  7 
[4]  0  8  4  9  2 10  5 11  1 12  6(13) 3  7 
[5]  0  8  4  9  2 10  5 11  1 12  6 13  3(14) 7 
[6]  0  8  4  9  2 10  5 11  1 12  6 13  3 14  7(15)
[7]  0(16) 8  4  9  2 10  5 11  1 12  6 13  3 14  7 15 
[8]  0 16  8(17) 4  9  2 10  5 11  1 12  6 13  3 14  7 15 
[9]  0 16  8 17  4(18) 9  2 10  5 11  1 12  6 13  3 14  7 15 
[1]  0 16  8 17  4 18  9(19) 2 10  5 11  1 12  6 13  3 14  7 15 
[2]  0 16  8 17  4 18  9 19  2(20)10  5 11  1 12  6 13  3 14  7 15 
[3]  0 16  8 17  4 18  9 19  2 20 10(21) 5 11  1 12  6 13  3 14  7 15 
[4]  0 16  8 17  4 18  9 19  2 20 10 21  5(22)11  1 12  6 13  3 14  7 15 
[5]  0 16  8 17  4 18(19) 2 20 10 21  5 22 11  1 12  6 13  3 14  7 15 
[6]  0 16  8 17  4 18 19  2(24)20 10 21  5 22 11  1 12  6 13  3 14  7 15 
[7]  0 16  8 17  4 18 19  2 24 20(25)10 21  5 22 11  1 12  6 13  3 14  7 15
```

The goal is to be the player with the highest score after the last marble is used up. Assuming the example above ends after the marble numbered 25, the winning score is 23+9=32 (because player 5 kept marble 23 and removed marble 9, while no other player got any points in this very short example game).

Here are a few more examples:

`10` players; last marble is worth `1618` points: high score is `8317`
`13` players; last marble is worth `7999` points: high score is `146373`
`17` players; last marble is worth `1104` points: high score is `2764`
`21` players; last marble is worth `6111` points: high score is `54718`
`30` players; last marble is worth `5807` points: high score is `37305`


*What is the winning Elf's score?*


```{r input}
parse_input <- function(.s) {
  
  str_split(.s, "(?i)players; last marble is worth") |>
    pluck(1) |> 
    parse_number() |> 
    as.integer()
  
}

input <- read_lines("input data/day09_input.txt") |> parse_input()
test1 <- "9 players; last marble is worth 25 points" |> parse_input()
test2 <- "10 players; last marble is worth 1618 points" |> parse_input()
test3 <- "13 players; last marble is worth 7999 points" |> parse_input()
test4 <- "17 players; last marble is worth 1104 points" |> parse_input()
test5 <- "21 players; last marble is worth 6111 points" |> parse_input()
test6 <- "30 players; last marble is worth 5807 points" |> parse_input()

```


```{r puzzle1, eval=FALSE}
run_game <- function(.in = input, puzzle2 = FALSE) {
  
  n_players <- .in[[1]]
  max_value <- .in[[2]]
  
  if (puzzle2 == TRUE) {max_value <- max_value * 100}
  
  marbles_master <- rep(NA_real_, max_value)
  ball_value <- 0L
  current_marble <- 1L
  player <- 0L
  player_scores <- vector("double", length = n_players)
  
  
  
  prep_next <- 
    function(.n = n_players, 
             .player = player, 
             .bv = ball_value
    ) {
      
      ball_value <<- .bv + 1L
      
      .p <- (.player + 1L) %% .n
      player <<- if_else(.p == 0L, .n, .p)
      
    }
  
  
  
  place_marble <- 
    function(.m = marbles, 
             .cm = current_marble, 
             .bv = ball_value, 
             .p = player, 
             .pscores = player_scores
    ) {
      
      # process for if a marble is divisible by 23...
      if (.bv %% 23L == 0L) {
        
        # figure out second marble to add to score
        .other_index <- .cm - 7
        .other_index <- if_else(.other_index < 1L,  
                                length(.m) + .other_index, 
                                .other_index)
        
        player_scores[[.p]] <<- .pscores[[.p]] + .bv + .m[[.other_index]]
        
        if (.other_index == 1L) {
          .mm <- c(.m[2:length(.m)], 
                   NA_integer_) # need a NA to replace the one we've removed
          
        } else {
          .mm <- c(.m[1:(.other_index - 1L)], 
                   .m[(.other_index + 1L):length(.m)], 
                   NA_integer_)
          
        }
        
        marbles_master[1:length(.mm)] <<- .mm
        current_marble <<- .other_index
        
        
        
        # process for every other marble than those divisible by 23
      } else {
        
        if (.cm == length(.m) ) {
          place <- 2L
          
        } else {
          place <- .cm + 2L
          
        }
        
        
        if (place == length(.m) + 1L) {
          .mm <- c(.m, .bv) 
          
        } else {
          .mm <- c(.m[1:(place - 1)], .bv, .m[place:length(.m)] )
          
        }
        
        
        marbles_master[1:length(.mm)] <<- .mm
        current_marble <<- place
      }
      
    } # end place_marble() function
  
  
  
  
  
  while (ball_value <= max_value) {
    
    marbles <- marbles_master[!is.na(marbles_master)]
    
    if (length(marbles) == 0L) {
      
      marbles_master[1] <- ball_value
      
      current_marble <- 1L
      
      prep_next()
      
      next
    }
    
    if (length(marbles) == 1L) {
      
      marbles_master[2] <- ball_value
      
      current_marble <- 2L
      
      prep_next()
      
      next
    }
    
    
    place_marble()
    
    prep_next()
    
    print(ball_value / (max_value * 1.0 ))
    
  }
  
  
  max(player_scores)
  
}

run_game()
```

`388844` is the score of the winning player.


First attempt at puzzle1 gets the right answer, but is inefficient and cannot complete puzzle2 in a reasonable time. Rebuilding code to hopefully run faster. 



```{r puzzle1 attempt 6, eval = TRUE}
# .in <- test6
# .puz2 <- FALSE
run_sim <- function(.in = input, .puz2 = FALSE) {
  
  max_players <- .in[[1]]
  max_marbles <- .in[[2]]
  
  if (.puz2 == TRUE) {
    max_marbles <- max_marbles * 100L
  }
  
  # initial conditions
  # player <- 1
  player_scores <- vector("double", length = max_players)
  
  current_ball <- 1L # R starts indexing at 1, so we will as well
  ball_pointer <- 1L # first index
  
  marbles <- rep(NA_integer_, max_marbles + 1)
  
  scoring_marbles <- vector("integer", length = (max_marbles %/% 23) + 1)
  scoring_marbles <- (23L * 1:((max_marbles %/% 23L) + 10L ) ) + 1L
  
  
  # function to run at end of each loop to prep the next execution
  #    advance player to next one
  #    update ball value top place by one
  # prep_next <- function(.p = player, 
  #                       .max = as.double(max_players), 
  #                       .iterate) {
  #   .p <- (.p + 1) %% .max
  #   .p <- if_else(.p == 0, .max, .p)
  #   
  #   player <<- .p
  #   current_ball <<- current_ball + 1L
  # }
  
  prep_next <- function(.iterate) {
    current_ball <<- current_ball + 1L
  }
  
  
  # function to run each loop 
  place_marble <- function(.bv = current_ball ) {
    
    if (.bv == min(scoring_marbles, na.rm = TRUE) || .bv %in% 1:2) {
      
      # for first ball, it will point to itself
      if (.bv == 1L) {
        marbles[1] <<- 1L
        ball_pointer <<- .bv
        
        prep_next()
        
        # for second ball, value "1", we should create our circular reference for the queue
      } else if (.bv == 2L) {
        marbles[1] <<- 2L
        marbles[2] <<- 1L
        ball_pointer <<- .bv
        
        prep_next()
        
        # logic for whenever ball value is divisible by 23 to add to player scores
        #    current ball value is not added to queue, but is added to the current player's score
        #    ball 7 to the left is removed from queue and also added to the current player's score
        #    references then updated to account for the removed ball
        #        ball 8 left points to ball 6 left
        #    current ball should then be the ball 6 to the left
      } else if (.bv == min(scoring_marbles, na.rm = TRUE)) {
        
        .left <- vector("integer", length = 8)
        .v <- ball_pointer
        for (i in 1:8) {
          .left[[i]] <- which(marbles == .v)
          .v <- .left[[i]]
        }
        
        # figure out the player
        .p <- .bv %% max_players
        .p <- if_else(.p == 0, max_players, .p)
        
        # need to subtract one from ball values due to indexing starting at 1 in R
        player_scores[[.p]] <<- player_scores[[.p]] + (.bv - 1L) + (.left[[7]] - 1L)
        
        marbles[[.left[[7]]]] <<- NA_integer_
        
        marbles[.left[8]] <<- .left[[6]]
        
        ball_pointer <<- .left[[6]]
        
        scoring_marbles[[which(scoring_marbles == min(scoring_marbles, na.rm = TRUE))]] <<- NA_integer_
        
        prep_next()
        
        print(current_ball / max_marbles)
      } 
      
        
        # normal ball placement and adjustment of queue references
        #    ball gets placed between balls 1 and 2 to the right
        #    ball one right will point to inserted ball
        #    inserted ball will point to ball two right
    } else if (length(scoring_marbles[is.na(scoring_marbles)]) < 2) {
      
      .one <- marbles[[ball_pointer]]
      .two <- marbles[[.one]]
      
      marbles[[.one]] <<- .bv
      marbles[[.bv]] <<- .two
      
      ball_pointer <<- .bv
      
      prep_next()
      
      # normal ball placement in chunks of 22 at a time
    } else {
      
      .inserts <- current_ball + 0:21
      
      .right <- vector("integer", length = 23)
      .v <- ball_pointer
      for (i in 1:23) {
        .right[[i]] <- marbles[[.v]]
        .v <- .right[[i]]
      }
      
      marbles[.right[1:22]] <<- .inserts
      marbles[.inserts] <<- .right[2:23]
      
      ball_pointer <<- .inserts[[22]]
      
      purrr::map(.x = 1:22, 
                 .f = ~ prep_next(.iterate = .x))
      
    }
    
  }
  
  
  while (current_ball <= max_marbles) {
    
    place_marble()
    
  }
  
  
  max(player_scores)
}


tictoc::tic("puzzle 1")

run_sim(.in = input, .puz2 = FALSE)

tictoc::toc()
```

## Puzzle 2

Amused by the speed of your answer, the Elves are curious:

*What would the new winning Elf's score be if the number of the last marble were 100 times larger?*


```{r puzzle2}
tictoc::tic("puzzle 2")

run_sim(.in = input, .puz2 = TRUE)

tictoc::toc()
```

`3212081616` is the score of the winning player if the number of the last marble is 100 times larger. 
